### 2020-06-23 chapters/background.tex
EMAs Pedro feedback applied

We talked about sliding window aggregations and algorithms to compute them and observe that the contents of the sliding window must be stored in memory so that when a new element is inserted we can evict the oldest one and discount its effect from the aggregation value. In this Section we analyze Exponential Moving Averages (EMAs). EMAs process each event only upon insertion. Since there is no eviction operation to perform, we do not need to store the whole window of data.

Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are applied to a sequence of elements to compute a weighted average, using different weights for each element. These weights decrease exponentially with the age of the observations. In other words, EMAs give higher weights to more recent data points.

\textcolor{red}{TODO: Images for EMA and sliding window weights}

@@

In this Thesis we will work with tuple-based EMAs but our approach can be adjusted to use time-based ones.

>>>>

We talked about sliding window aggregations and algorithms to compute them. We observed that the contents of the sliding window must be stored in memory in order to evict the oldest event and discount its effect from the aggregation when a new one is inserted. In this Section we analyze Exponential Moving Averages (EMAs). EMAs process each event only upon insertion. Since there is no eviction operation to perform, we do not need to store the whole window of data.

Exponential Moving Averages (also known as Exponential Weighted Moving Averages) \cite{EMA-Everett2011} \cite{EMA-Hunter} \cite{EMA-MarcusB} are a type of reduce functions applied to a sequence of elements, one at a time in a orderly fashion, to compute a weighted average, using different weights for each element. These weights decrease exponentially with the age of the observations. In other words, EMAs give higher weights to more recent data points. Figures \ref{fig:window-weights} and \ref{fig:ema-weights} show the weights applied to past events. In a typical sliding window, events either have weight of one (inside the window) or zero (outside the window) whereas EMAs give exponentially smaller weights to older events.

\begin{figure}[!htb]
\centering
\begin{minipage}{.5\textwidth}
  \centering
      \includegraphics[scale=0.5]{figures/window-weights.png}
  \captionof{figure}{Window weights}
  \label{fig:window-weights}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
      \includegraphics[scale=0.5]{figures/ema-weights.png}
  \captionof{figure}{EMA weights}
  \label{fig:ema-weights}
\end{minipage}
\end{figure}

@@

Definition \ref{def:ema} introduces the recurrence formula for time-based EMAs whereas Definition \ref{def:tuple-ema} works for tuple-based ones. In this Thesis we chose to work with tuple-based EMAs due to its simplicity but our approach can be adjusted to use time-based ones.

### 2020-06-23 chapters/my-work.tex
EMAs Pedro feedback applied

\textcolor{red}{NOTE TO MOPS: You commented here, but arent EMAs senseitive to time correlations? Since we put lower weights on older events, the order matters (while in a sliding window every event has the same weight (1))}
Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}. Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.

@@

A data stream is a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are time-dependent variables named features. They can be thought of as columns in a traditional relational database table. Our goal is to detect data pattern shifts for each feature in an online fashion. Such shifts will be measured between a reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

We do not need to directly compare the entire window contents of the reference and sliding windows. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} constant in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

Histograms are a representation of the distribution of numerical or categorical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require the storage of a fixed-size list of bins and a corresponding fixed-size list of counts per bin for ever growing volumes of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and adding or subtracting one, respectively, thus satisfying property \textit{(b)}. However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

@@

To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein. A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}. Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations. Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).

@@

\begin{enumerate}
    \item obtain a reference histogram for each feature using all data in the reference period

    \item then compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
\end{enumerate}

@@

It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself. For instance, given the JSD value varies between 0 and 1, why not just set to alert if the measured JSD value is 0.7? The reason is that a divergence value may fall on different percentiles for different features. In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features. Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.

@@

\subsubsection*{Batch phase artifacts}
For each feature, by the end of the batch phase, we have a reference histogram, a list of distance values that represents its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.

@@

For each feature, we initialize the histogram we will approximately maintain with the last sample's data, from the batch phase. Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}. This is all we do, regarding event processing.

Periodically, we perform a multi-feature change detection test. The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of feature \textit{x}. This distance value is then tested against the known distribution of distance values for feature \textit{x}. In other words, we test to see the percentile of the computed distance in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile. When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations. The probability value or \textit{p-value} of a distance value is

@@

Hence, if we set to alert states that have 1\% or less probability to happen, we raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01. Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:

@@

\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis. When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives. In other words, when repeating a test multiple times the multiple test problem arises which states that the more inferences are made, the more likely erroneous inferences are to occur. In the literature, many multiple test correction methods are proposed, but we opt to use the Holm-Bonferroni multiple test correction. The Holm-Bonferroni method is one of many approaches for controlling the so called Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.

@@

We are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests. We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$. In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.

>>>>

\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
Both time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}. Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.

@@

A data stream is a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are time-dependent variables named features. They can be thought of as columns in a traditional relational database table. Our goal is to detect data pattern shifts for each feature in an streaming fashion. Such shifts will be measured between a reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

We do not need to directly compare the entire window contents of the reference and sliding windows. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

One aggregation that encodes distributions properly are histograms. Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require storing a fixed-size list of bins and corresponding counts per bin for ever growing volumes of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}. However, computing an histogram aggregation over a sliding window of size $W$ implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

@@

To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein. A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}. Given a divergence value for the two aggregations (reference and sliding) we need a measure of normal statistical fluctuations. Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).

@@

\begin{enumerate}
    \item Obtain a reference histogram for each feature using all data in the reference period

    \item Compute the distribution of distances between the reference histogram and the sliding window histogram by computing the distance (or divergence value) between the reference histogram and various samples of sliding window histograms
\end{enumerate}

@@

It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself. For instance, given that JSD values varie between 0 and 1, why not just set to alert if the measured JSD value is 0.7? The reason is that a divergence value may fall on different percentiles for different features. In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features. Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.

@@

\subsubsection*{Batch phase artifacts}
By the end of the batch phase, we should have, for each feature: A reference histogram, a list of distance values that represent its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.

@@

For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase. Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}. This is all we do, regarding event processing.

In order to obtain alerts, we periodically perform a multi-feature change detection test. The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of the same feature. This distance value is then tested against the known distribution of distance values for feature \textit{x}. In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile. When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations. The probability value or \textit{p-value} of a distance value is

@@

For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01. Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:

@@

\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis. When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives. In other words, when repeating a test multiple times the multiple test problem arises which states that the more inferences are made, the more likely erroneous inferences are to occur. In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction. The Holm-Bonferroni method is one of many approaches for controlling the so called Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.

@@

When checking for alerts we are performing one test for each feature, hence we performing \textit{n\textsubscript{features}} tests. We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$. In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.
